---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# `naturecounts`

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)

Access and download data on plant and animal populations from various databases through NatureCounts, a service managed by Bird Studies Canada. 

## Installation

You can install this developmental version of `naturecounts` from GitHub with the remotes package:

```{r eval = FALSE}
install.packages("remotes")
install_github("BirdStudiesCanada/naturecounts")
```

## Usage

```{r}
library(naturecounts)
```


### Fetching data counts

Use the `nc_count()` function to return all public collections and the number of observations in each.

```{r}
nc_count()
```

If you have a [NatureCounts username](https://www.birdscanada.org/birdmon/default/register.jsp) you can supply it with the `username` argument (you will be prompted for a password). `nc_count()` will now return all the records you have access to.

```{r, eval = FALSE}
nc_count(username = "my_user_name")
```

### Fetching data

Fetch all observations of moose which are publically available to a local data frame.
```{r}
moose <- nc_data_dl(species = 133990)
```

Alternatively, save the downloaded data as a SQLite database (`moose.nc`).

```{r}
moose <- nc_data_dl(species = 133990, sql_db = "moose")
```

```{r cleanup, include = FALSE}
file.remove("moose.nc")
```


> See the article on [Species Codes](articles/species-codes.html) to determine which codes to use.

### Metadata

NatureCounts includes a great deal of metadata which can be accessed through the functions with the `meta_` prefix. See the [Meta Documentation](reference/meta.html) for specifics.
