---
title: "Chapter 3 - Understanding and Viewing Data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Understanding and Viewing Data {#Data3}
This chapter begins with a brief introduction to the structure of the NatureCounts database, followed by a description of access levels and how to create a user account. We then provide instructions on how to view data from various collections and apply filters.

```{r}
library(naturecounts)
```

> The code in this Chapter will not work unless you replace `"testuser"` with your actual user name. You will be prompted to enter your password.

## Data Structure {#Data3.1}

The [Bird Monitoring Data Exchange](https://naturecounts.ca/nc/default/nc_bmde.jsp) (BMDE) was developed to be a standardized data exchange schema to promote the sharing and analysis of avian observational data. The schema is the core sharing standard of the [Avian Knowledge Network](http://www.avianknowledge.net).The BMDE (currently version 2.0) includes 169 *core* fields (variables) that are capable of capturing all metrics and descriptors associated with a bird observation. The BMDE schema was extended in 2018, and the *complete* version now includes 265 fields (variables).

> **Fields** are variables or columns in a data set

By default, the naturecounts package downloads the data with the *minimum* set of fields/columns. However, for more advanced applications, users may wish to specify which fields/columns to return using the `field_set` and `fields` options in the `nc_data_dl()` function. For help with this feature, see the naturecounts article ['Selecting columns and fields to download'](https://birdscanada.github.io/naturecounts/articles/selecting-fields.html).

## Levels of Data Access {#Data3.2}

NatureCounts hosts many datasets, representing in excess of 170 million occurrence records, with a primary focus on Canadian bird monitoring data. Many of those datasets are from projects lead by Birds Canada and/or its partners. While we strive to make our data as openly available as possible, we also need to recognize the needs of our partners and funders.

NatureCounts has five [Levels of Data Access](https://naturecounts.ca/nc/default/nc_access_levels.jsp), which define how each dataset can be used. Those levels are set individually for each dataset, in consultation with the various partners and data custodians involved.

-   Level 0: most restricted (archival only)

-   Level 1: archival only, metadata visible

-   Level 2: data used for visualizations only

-   Level 3: data available to third parties by request

-   Level 4: data shared with external portals and available by request

-   Level 5: open access

All contributing members of NatureCounts have complete authority over the use of the data they have provided, and can withhold data at any time from any party or application. All users of any NatureCounts data must clearly acknowledge the contribution of the members who are making data available. Each dataset comes with its own [Data Sharing Policy](https://naturecounts.ca/nc/default/nc_data_sharing.jsp) that defines the various conditions for data usage.

You can view the Data Access Level for each collection on the [NatureCounts Datasets](https://naturecounts.ca/nc/default/datasets.jsp) page or using the [metadata](#Data3.5) function (see `akn_level`):

```{r access, eval = FALSE}
collections<-meta_collections()
View(collections)
```

> You can create a stand alone table for any metadata table using similar syntax as above

To retrieve the metadata for a specific set of collections based on its project_id, you can index the 'collections' dataframe.

```{r project id, eval=FALSE}
# Specify the project_id you want to retrieve
projectID <- "insert_project_id_number"

# Subset to retrieve the relevant metadata based on project_id
collection_metadata <- collections[collections$project_id == projectID, ]

# View the metadata of the collection(s) with the specified project_id
if (nrow(collection_metadata) > 0) {
  View(collection_metadata)
} else {
  print("No collection found with the specified project_id.")
}
```

## Authorizations {#Data3.3}

To access data using the naturecounts R package, you must [sign up](https://naturecounts.ca/nc/default/register.jsp) for a **free** account. Further, if you would like to access Level 3 or 4 collections you must make a [data request](https://naturecounts.ca/nc/default/searchquery.jsp). For step-by-step visual instructions, we encourage you to watch: [NatureCounts: An Introductory Tutorial](link%20to%20be%20provided).

> Create your **free** account now before continuing with this workbook

## Viewing information about NatureCounts collections {#Data3.4}

First, lets use the naturecounts R package to view the number of records available for different collections. To do this we use the `nc_count()` function. You can view *all* the available collections and the number of observations using the default setting.

If a username is provided, the collections are filtered to only those available to the user. Otherwise all counts from all data sources are returned (default: `show = "all"`).

```{r view all, eval=FALSE}
nc_count <- nc_count()
View(nc_count)
```

Or you can view the collections for which you have access using your username/password.

```{r view username, eval=FALSE}
nc_count <- nc_count(username = "testuser")
View(nc_count)
```

Further refinements can be applied to the `nc_count()` function using [filters](#Download4) Options include: `collections`, `project_id`, `species`, `years`, `doy` (day-of-year), `region`, and `site_type`.

## Metadata codes and descriptions {#Data3.5}

There are [metadata](https://birdstudiescanada.github.io/naturecounts/reference/meta.html) associated with the various arguments used in the `nc_count()` and `nc_data_dl()` functions, the latter you will use in [Chapter 4](#Download4). These are stored locally and can be accessed anytime to help filter your data view or download query. They include:

-   `meta_country_codes()`: country codes

-   `meta_statprov_codes()`: state/Province codes

-   `meta_subnational2_codes()`: subnational2 codes

-   `meta_iba_codes()`: Important Bird Area (IBA) codes

-   `meta_bcr_codes()`: Bird Conservation Region (BCR) codes

-   `meta_utm_squares()`: UTM Square codes

-   `meta_species_authority()`: species taxonomic authorities

-   `meta_species_codes()`: alpha-numeric codes for avian species

-   `meta_species_taxonomy()`: codes and taxonomic information for all species

-   `meta_collections()`: collections names and descriptions

-   `meta_breeding_codes()`: breeding codes and descriptions

-   `meta_project_protocols()`: project protocols

-   `meta_projects()`: projects ids, names, websites, and descriptions

-   `meta_protocol_types()`: protocol types and descriptions

Any of these functions can be used to browse the code lists relevant to your search. For example, you can view the metadata for Birds Canada projects, including project ids using:

```{r meta, eval = FALSE}
project_ids <- meta_projects() # retrieve the project_ids represented in the repository 
View(project_ids) # explore the dataframe 
```

Using the above functions to search by country, state/province, subnational2, BCR etc. is especially useful for regional filtering in this next section.

## Region & Species filtering {#Data3.6}

Filtering will often be done based on geographic extent (i.e., `region`). To filter by `region` you must provide a named list with *one* of the following:

-   `country`: country code (e.g., CA for Canada)

-   `statprov`: state/province code (e.g., MB for Manitoba)

-   `subnational2`: subnational (type 2) code (e.g., CA.MB.07 for the Brandon Area)

-   `iba`: Important Bird Areas (IBA) code (e.g., AB001 for Beaverhill Lake in Alberta)

-   `bcr`: Bird Conservation Regions (e.g., 2 for Western Alaska)

-   `utm_squares`: UTM square code (e.g., 10UFE96 for a grid in Alberta)

-   `bbox`: bounding box coordinates (e.g., c(left = -101.097223, bottom = 50.494717, right = -99.511239, top = 51.027557) for a box containing Riding Mountain National Park in Manitoba). On the NatureCounts web portal there is a handy [Within Coordinates](https://www.birdscanada.org/birdmon/default/searchquery.jsp)) tool to help you retrieve custom coordinates for your data query and/or download.

The search_region function may be used to search a region by name (English or French) by specifying the 'type' argument.

For example, by country;

```{r}
search_region("États-Unis", type = "country")
```

Or by Bird Conservation Region:

```{r}
search_region("rainforest", type = "bcr")
```

When using the `nc_count()` view function, you have the helpful option of filtering data by region. Let's demonstrate how to use this function and the `region` argument with a few examples.

First, let's limit our data search to Quebec:

```{r nc_count example, eval=FALSE}
nc_count(region = list(statprov="QC")) # filter nc_count by statprov
```

Next, let's say we want to narrow down our search to the subnational level (Montreal and Toronto) but don't know the corresponding codes for these regions.

Browse the code list:

```{r}
View(meta_subnational2_codes())
```

Or, more efficiently, search by region:

```{r}
search_region("Montreal", type = "subnational2")
```

Great, we now know the codes we need and can view our metadata using `nc_count()`:

```{r}
nc_count(region = list(subnational2 = c("CA-QC-MR", "CA-ON-TO")))
```

Similarly, this function can be used to view metadata for a bounding box using latitude and longitude coordinates:

```{r}
nc_count(region = list(bbox = c(left = -125, bottom = 45, 
                                right = -100, top = 50)))
```

Another commonly used filter is specific to `species`. In order to filter by `species` you need to get the species id codes. These are numeric codes that reflect species identity.

For all species, you can search the NatureCounts repository by scientific, English or French name with the search_species() function.

```{r species example, eval=FALSE}
search_species("chickadee") # returns all chickadee species ids
```

The corresponding species id can then be used to download the data either directly or by saving and referencing the data frame: [see Chapter 4](#Data4)

For birds, you can also search by alphanumeric species code with the search_species_code() function.

```{r}
species_codes <- search_species_code() # returns all species codes represented in the database.
```

This function, by default, uses the BSCDATA taxonomic authority and returns all species codes related to the search term including related subspecies. For this reason, it is considered a more robust method for ensuring that you do not miss observations in your search.

If you're interested in a particular species, and any recognized subspecies, you can filter your search:

```{r species code search example, eval=FALSE}
search_species_code("BCCH") # Filters by species code
```

The search function is case insensitive:

```{r}
search_species_code("bcch") # Also filters by species code
```

It is important to note that species subdivisions (subspecies, subpopulations, hybrids, etc.) can also be recognized with different codes across taxonomic authority (BSCDATA, CBC).

For example, BSCDATA recognizes 3 sub groups of the Dark-eyed junco:

```{r}
search_species_code("DEJU") # Returns species ids for Junco hyemali and 3 related subgroups
```

We have to modify our search when filtering by the CBC code system. This taxonomic authority recognizes 9 sub group hybrids of the Dark-eyed junco, as well as the Guadalupe junco:

```{r}
search_species_code("12385", authority = "CBC") # Returns the species ids and CBC codes for Junco hyemali, 9 subgroups, and the Guadelupe junco
```

You can search by more than one authority at the same time. Note that your search term only needs to match one authority (not both), and that the information returned reflects both authorities combined.

```{r}
View(search_species_code("DEJU", authority = c("BSCDATA", "CBC")))
```

If you do not want all subgroups, you can use the results = "exact" argument to return only an exact match.

```{r}
search_species_code("DEJU", results = "exact")
```

For additional examples and more advanced options are available online for retrieving [Region](https://birdstudiescanada.github.io/naturecounts/articles/region-codes.html) and [Species](https://birdscanada.github.io/naturecounts/articles/species-codes.html#advanced-searches) codes.

## Examples {#Data3.7}

Here are a few examples for you to work through to become familiar with the `nc_count()` function.

*Example 1*: Determine the number of collections and records for a specific *region*. The options include: `country`, `statprov`, `subnational2`, `iba`, `bcr`, `utm_squares`, and `bbox`. You can find details and examples on how to [`search_region()`](https://birdstudiescanada.github.io/naturecounts/articles/region-codes.html) at the link provided.

The following code will retrieve all available collections and number of records for British Columbia

```{r nc_region, warning = FALSE, eval=FALSE}
search_region("British Columbia", type = "statprov")
nc_count(region = list(statprov = "BC"))
```

*Example 2*: Determine the number of records for a specific *species*. You can find details and examples on how to [`search_species_code()`](https://birdstudiescanada.github.io/naturecounts/reference/search_species_code.html) based on 4 letter alpha code and [`search_species()`](https://birdstudiescanada.github.io/naturecounts/reference/search_species.html) based on common names at the links provided.

The following code will retrieve all available collections and number of records for Red-headed Woodpecker

```{r nc_species1, eval=FALSE}
search_species("Red-headed Woodpecker")
search_species_code("RHWO")

RHWO<-nc_count(species = 10060)
View(RHWO)
```

*Example 3*: We can further refine the Red-headed Woodpecker example (above) by filtering the species-specific data by region (e.g., [Bird Conservation Region](http://nabci-us.org/assets/images/bcr_map2.jpg) 11), time period (e.g., 2015-2019), or a combination of both.

```{r nc_species2, eval=FALSE}
RHWO_11 <- nc_count(species = 10060, region = list(bcr = "11"))
View(RHWO_11)

RHWO_year <- nc_count(species = 10060, year = c(2015, 2019))
View(RHWO_year)

RHWO_11_year <- nc_count(species = 10060, region = list(bcr = "11"), year = c(2015, 2019))
View(RHWO_11_year)
```

## Exercises {#Data3.8}

Now apply your newly acquired skills!

*Exercise 1*: If you are interesting in doing a research project on Snowy Owls in Quebec, which three collections are you most likely to consider using (i.e., which have the most data)?

Answer: EBird-CA-QC, OISEAUXQC, CBC

*Exercise 2*: How many records of Gadwal are in the [British Columbia Coastal Waterbird Survey](https://www.birdscanada.org/birdmon/atowls/datasets.jsp?code=BCCWS) collection? What if you are only interested in records from 2010-2019, how many records are available?

Answer: 702, 389

Full answers to the exercises can be found in [Chapter 7](#Ans7.1).
