---
title: "COSEWIC Tools"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{COSEWIC Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message = FALSE}
library(naturecounts)
```


Basic plot
```{r}
r <- cosewic_ranges(bcch)
cosewic_plot(r, title = "Black-capped Chickadee")
```

Adding observation points
```{r}
cosewic_plot(r, points = bcch, title = "Black-capped Chickadee")
```

Only EOO or IAO

```{r}
cosewic_plot(r, which = "eoo", points = bcch)
cosewic_plot(r, which = "iao")
```

Change the CRS

- Only applies if not using map tiles as they *must* be in the CRS of the tile (i.e. EPSG:3857 Web Mercator)

```{r}
cosewic_plot(r, crs = 3347) # No change
cosewic_plot(r, map = map_canada(), crs = 3347)
```

Move the scale/arrow

```{r}
r <- cosewic_ranges(hofi)
cosewic_plot(r, arrow_location = "br", scale_location = "br")
```

Summarize IAO over larger grid for better visibility
```{r}
cosewic_plot(
  r,
  grid = grid_canada(25),
  title = "House Finch",
  arrow_location = "br",
  scale_location = "br"
)
```

Plot multiple species as separate plots
```{r}
m <- rbind(bcch, hofi)
r <- cosewic_ranges(m)
p <- cosewic_plot(
  r,
  title = c("14280" = "Black-capped chickadees", "20350" = "House Finches")
)
p[[1]]
p[[2]]
```

Use patchwork to combine
```{r}
#| fig-width: 10
#| fig-asp: 0.5
library(patchwork)
wrap_plots(p) +
  plot_layout(guides = "collect")
```

Use IAO as a proportion for better legends

```{r}
#| fig-width: 10
#| fig-asp: 0.5
p <- cosewic_plot(
  r,
  iao_prop = TRUE,
  title = c("14280" = "Black-capped chickadees", "20350" = "House Finches")
)

wrap_plots(p) +
  plot_layout(guides = "collect")
```

Consider summarizing over a larger IAO grid (10x10km) for better visibility
```{r}
#| fig-width: 10
#| fig-asp: 0.5
p <- cosewic_plot(
  r,
  iao_prop = TRUE,
  grid = grid_canada(10),
  title = c("14280" = "Black-capped chickadees", "20350" = "House Finches")
)

wrap_plots(p) +
  plot_layout(guides = "collect")
```

For more nuanced control, create plots separately and then combine


```{r}
#| fig-width: 10
#| fig-asp: 0.5
b <- cosewic_ranges(bcch)
h <- cosewic_ranges(hofi)
p1 <- cosewic_plot(b, title = "Black-capped chickadee", iao_prop = TRUE)
p2 <- cosewic_plot(
  h,
  title = "House Finches",
  iao_prop = TRUE,
  arrow_location = "br",
  scale_location = "br",
  grid = grid_canada(25)
)

p1 + p2 + plot_layout(guides = "collect")
```

```{r}
#| fig-asp: 1.5

p1 / p2 + plot_layout(guides = "collect")
```