---
title: "COSEWIC Tools"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{COSEWIC Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options("pillar.width" = 120)
```

```{r, message = FALSE}
library(naturecounts)
library(dplyr) # For manipulating data frames
library(patchwork) # For combining plots
library(ggplot2) # For plotting the grid
```

## Getting started

These tools, `cosewic_ranges()` and `cosewic_plot()`, are designed to help with
spatial calculations for COSEWIC assessments, namely calculations of the EOO and IAO (see`cosewic_ranges()` for more details on these calculations).

You can calculate both IAO and EOO with the `cosewic_ranges()` function and you can use the `cosewic_plot()` function to create figures of these values.

In the next few examples, we'll use a built in dataset `bcch`.
In your own workflows, you would replace this with your own data (see [Using your own data](#using-your-own-data)).

### Calculating IAO and EOO

First we'll calculate the ranges using default arguments and call `r`.

```{r}
#| message: false
r <- cosewic_ranges(bcch)
```

Look at this data by printing the `r` object.
This shows us that the `r` object is a list with two items:

- `iao` which is a simple features collection (sf or spatial dataframe), and 
- `eoo` which is also an sf or spatial data frame.

```{r}
r
```

You can access either of these items with the `$` to pull out just what you're interested in.

```{r}
r$eoo
```

The values you are likely to be especially interested in are the `iao` and the `eoo` columns within these spatial dataframes

```{r}
r$iao$iao[1]
r$eoo$eoo[1]
```

Also of interest is the `prop_include` column in each set of data. 
This remindes you of the proportion of original data points retained in the calculations (i.e. `prop_include = 1` is 100%).
You can change the proportion of points included to omit outliers if you like by modifiying the `prop_include` argument in `cosewic_ranges()`.

By default all points are included, so make sure you're confident that those points are accurate!

If this is too much information, omit the spatial data from the range calculations.

```{r}
cosewic_ranges(bcch, spatial = FALSE)
```

### Plotting IAO and EOO
By default, `cosewic_ranges()` includes the spatial data so we can easily create figures of these ranges.

```{r}
r <- cosewic_ranges(bcch)
cosewic_plot(r, title = "Black-capped Chickadee")
```

> **Remember**: By default we use map tiles from OpenStreetMap (`map = "osm"`). 
> If you are using these figures in a public document/website/etc.,
> you must [attribute OpenStreetMap](https://osmfoundation.org/wiki/Licence/Attribution_Guidelines).

You can try any map tile listed in `rosm::osm.types()`, but note that not all may work for your region and many require an API key.

```{r}
r <- cosewic_ranges(bcch)
cosewic_plot(r, map = "cartolight", title = "Black-capped Chickadee")
```

## Using your own data

To use your own data in the `cosewic_ranges()` function, you must load a data set of 
observations into R. This dataset must have an ID column and columns defining
latitude and longitude, and optionally, a grouping column. 

Let's load the example black-capped chickadee file included in naturecounts.
We'll use the `system.file()` function to find the path to the csv file.

```{r}
# Assign the path or location
path <- system.file("extdata", "bcch.csv", package = "naturecounts")
path
```

For your data, you'll a path something like `path <- "location/of/my/data.csv"` depending where your data is stored. See the [R for Data Science chapter on Data import](https://r4ds.hadley.nz/data-import.html#reading-data-from-a-file) for more details.

Now we'll read the data and take a quick look.
```{r}
bc <- read.csv(path)
head(bc)
```

To use this in `cosewic_ranges()` we'll tell the function how to interpret these columns. 
By default the function expects these columns to be `record_id`, 
`latitude` and `longitude`, so if they are different we need to specify what they are.
In our example we also specify `group = NULL` to tell the function that we're not using a grouping column.

```{r}
r <- cosewic_ranges(
  bc,
  record = "id",
  coord_lat = "lat",
  coord_lon = "lon",
  group = NULL
)
```



## Working with multiple DUs (Designatable Units)

Let's assume we have two populations or Designatable Units which we would like
to work with. We'll use the built in `pops` data set for this.

```{r}
head(pops)
```

Because we have multiple groups, we'll use the `group = "population"` option to 
tell the function which column contains the groups. Then calculations are performed separately for each group.

```{r}
r <- cosewic_ranges(pops, group = "population")
r$eoo
```

And we get a list of plots, one for each group.
```{r}
p <- cosewic_plot(r, group = "population")
p[[1]]
p[[2]]
```

For a combined plot, we can use the `wrap_plots()` function from the patchwork package to combine these figures.
```{r}
#| fig-width: 10
#| fig-asp: 0.3
wrap_plots(p)
```



## Using the COSEWIC's IAO grid

By default, `cosewic_ranges()` constructs an IAO grid for each analysis. 
This means that there might be small discrepancies in IAO values calculated by
naturecounts and those calculated with a different grid. 
Even if the grids are the same size and with the same projection, exactly where the bounds of the grid are may not be the same.

If this is of concern, you can supply your own IAO grid for these calculations, 
to ensure that there are no discrepancies.

First, use the sf package to read in your IAO grid. 
Here we'll use a mini example IAO grid stored in the package
```{r}
path <- system.file("extdata", "iao_bcch_grid.gpkg", package = "naturecounts")
path

grid <- sf::st_read(path)
```

This is what our example grid looks like
```{r}
ggplot(data = grid) + geom_sf()
```

Next we'll pass this as an argument to our `cosewic_ranges()` function.
```{r}
r <- cosewic_ranges(bcch, iao_grid = grid)
```

And take a look at the figure of these calculations.
```{r}
cosewic_plot(r)
```

For comparison, note how the calculations are very slightly different when using 
`cosewic_ranges()`'s default grid.
```{r}
r0 <- cosewic_ranges(bcch)
cosewic_plot(r0)
```

## Appendix

### Customizing Plots

These examples all use the `bcch` dataset.

```{r}
r <- cosewic_ranges(bcch)
```

#### Adding observation points
Note that this adds *all* observation points (regardless of whether some were omitted from the analyses).
```{r}
cosewic_plot(r, points = bcch)
```

#### Plot only either EOO or IAO

```{r}
cosewic_plot(r, which = "eoo", points = bcch)
cosewic_plot(r, which = "iao")
```

#### Change the CRS

Only applies if not using map tiles as map tile projections cannot be changed.

No change
```{r}
cosewic_plot(r, crs = 3347)
```

Using a custom polygon, we can change the CRS.
```{r}
cosewic_plot(r, map = map_canada(), crs = 3347)
```

#### Move the scale/arrow

```{r}
r <- cosewic_ranges(hofi)
cosewic_plot(r, arrow_location = "br", scale_location = "br")
```

#### Summarize IAO over larger grid

When the cells are really small, it can be helpful to summarize over a larger grid
for better visibility of the patterns.
```{r}
cosewic_plot(
  r,
  grid = grid_canada(25),
  title = "House Finch",
  arrow_location = "br",
  scale_location = "br"
)
```

### Customizing Multiple Plots

These examples all use the `pops` dataset.

```{r}
r <- cosewic_ranges(pops, group = "population")
```

#### Using IAO proportions

When plotting multiple DUs the IAO scales may be different enough that it would
be better use IAO proportions rather than absolute values. 

```{r}
#| fig-width: 10
#| fig-asp: 0.5
p <- cosewic_plot(r, group = "population", iao_prop = TRUE)

wrap_plots(p) +
  plot_layout(guides = "collect")
```

#### Summarize IAO over larger grid
As for single plots, when the cells are really small, it can be helpful to summarize over a larger grid
for better visibility of the patterns.

```{r}
#| fig-width: 10
#| fig-asp: 0.5
p <- cosewic_plot(
  r,
  group = "population",
  iao_prop = TRUE,
  grid = grid_canada(10)
)

wrap_plots(p) +
  plot_layout(guides = "collect")
```

#### Combining separate plots
But for the ultimate control, create plots separately and then combine.

We'll first split the two populations, but you could also load them separately 
if they are already split on your computer (see "Working with your own data").

We'll use `group = "population` to ensure correct titles.

```{r}
# Split the two populations
pops1 <- dplyr::filter(pops, population == "Population 1")
pops2 <- dplyr::filter(pops, population == "Population 2")

# Calculate the ranges separately
r1 <- cosewic_ranges(pops1, group = "population")
r2 <- cosewic_ranges(pops2, group = "population")

# Create separate plots
p1 <- cosewic_plot(r1, group = "population", iao_prop = TRUE)
p2 <- cosewic_plot(
  r2,
  group = "population",
  iao_prop = TRUE,
  grid = grid_canada(10)
)
```

Now arrange your plots how you like.

```{r}
#| fig-asp: 0.5
p1 +
  p2 +
  plot_layout(guides = "collect") +
  plot_annotation(title = "EOO and IAO for all Designatable Units")
```

```{r}
#| fig-asp: 1.5
p1 /
  p2 +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "EOO and IAO for all Designatable Units",
    subtitle = "2025 Assessment performed with naturecounts R package"
  )
```