---
title: "Data Access"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message = FALSE}
library(naturecounts)
```

## Understanding Data Access
All downloads in `naturecounts` require a free NatureCounts account ([sign up here](https://www.birdscanada.org/birdmon/default/register.jsp)). 

Not all data hosted by NatureCounts is immediately available to users. Different data sets have different access levels (also called `akn_levels`). A full description of these different levels is available from the [NatureCounts site](https://www.birdscanada.org/birdmon/default/nc_access_levels.jsp).

Briefly, there are five levels and the R package `naturecounts` can potentially access three of these levels (for most users).

Data at **access level 5** are considered public data and are available to all users. 
Data at **access levels 4 and 3** are private/semi-public projects/collections. To access this data you must request access (see Requesting Access, below).

Data at **access level 2** are visible in counts returned by `nc_count()` and in metadata returned by `meta_collections()` , etc., but are not downloadable through `nc_data_dl()` in the `naturecounts` R package without special permissions.

Data at **access level 1** are not visible through the `naturecounts` R package without special permissions.

## Requesting Access

You can request access to **full** or **partial** datasets at levels 3 and 4 through the NatureCounts [data request page](https://www.birdscanada.org/birdmon/default/searchquery.jsp). To request **partial** access, choose the filters and walk through the steps. To request **full** access, choose the collection and check the "Full access?" box in Step 2.

Once a request has been submitted it has to be approved by the appropriate data custodian. 

## What Do I Have Access To?

To see what you have access to you can use three functions: `nc_permissions()`, `nc_count()` and `nc_requests()`.

### Full Access - `nc_permissions()` or `nc_count()`
You can see which data you have **full** access to (i.e. you have access to all data within a specific collection) with either the `nc_permissions()` or `nc_count()` functions.

To get a list of collections to which you have access, use the `nc_permissions()` function.

```{r}
nc_permissions(username = "sample")
```

To additionally get the amount of data in each of these collections, user the `nc_count()` function.
```{r}
nc_count(username = "sample")
```

To see which data you do not (currently) have access to, use the `show = "all"` argument.
```{r}
nc_count(username = "sample", show = "all")
```

Data where `access` is "by request" is available through the online request portal. 
Data where `access` is "no access" may be available as counts, but is not generally accessible as a download of raw data.

### Partial Access - `nc_requests()`
You can see which data you have **partial** access to (i.e. you have requested access to part of a collection or part of a series of collections through the online portal) with the `nc_requests()` function. 

```{r, eval = FALSE}
nc_requests(username = "USER")
```

An example of the output you might see:
```
  request_id requestOrigin                     requestLabel      collection   status nrecords
1     155882           web                    steffi's test          QCOWLS  pending    12640
2     155886           web                 steffi's test SK   SKATLAS1BE_DO approved    99872
3     155886           web                 steffi's test SK SKATLAS1BE_SUMM approved    42203
4     155886           web                 steffi's test SK      SKATLAS1PC approved    27916
5     155888           web steffi's test partial projects 1     PFW-US-EAST  pending   368672
6     155888           web steffi's test partial projects 1     PFW-US-WEST  pending   282745
7     155889           web steffi's test partial projects 2     PFW-US-EAST  pending     1269
8     155889           web steffi's test partial projects 2     PFW-US-WEST  pending     4080
                                                                  filters
1                                       date_type (s); project ('QCOWLS')
2 date_type (s); project ('SKATLAS1BE_DO','SKATLAS1BE_SUMM','SKATLAS1PC')
3 date_type (s); project ('SKATLAS1BE_DO','SKATLAS1BE_SUMM','SKATLAS1PC')
4 date_type (s); project ('SKATLAS1BE_DO','SKATLAS1BE_SUMM','SKATLAS1PC')
5   date_type (s); project ('PFW-US-EAST','PFW-US-WEST'); species (14280)
6   date_type (s); project ('PFW-US-EAST','PFW-US-WEST'); species (14280)
7   date_type (s); project ('PFW-US-EAST','PFW-US-WEST'); species (14320)
8   date_type (s); project ('PFW-US-EAST','PFW-US-WEST'); species (14320)
```

The `status` field/column shows which requests have been `approved`, are `pending` or have been `declined`.

## How Do I Download Accessible Data?

### Full Access
Data to which you have full access can be directly downloaded via the `nc_data_dl()` function.

```{r, eval = FALSE}
moose <- nc_data_dl(species = 133990, username = "sample", info = "nc_example")
```

Note that `request_id`s are created for `web` requests (above) and are also created for `api` requests: data downloaded via the `nc_data_dl()` function if a `request_id` was not provided.

For example, the previous download created an `api` `request_id`. To see `request_id`s created through this R package, use the argument `type = "api"` in `nc_requests()`.
```{r}
nc_requests(username = "sample", type = "api")[1,]
```

This data can be re-downloaded (more quickly) by using this stored `request_id`.
```{r, eval = FALSE}
moose <- nc_data_dl(request_id = 152446, username = "sample")
```

Note that the "sample" user account no longer stores api requests, but you can still recover old requests as examples.

## How Do I Download Specially Requested Data?

Approved requests can either be downloaded by hand from the website, or the `request_id` can be used to download data with the `nc_data_dl()` function.

```{r, eval = FALSE}
nc_requests(username = "USER")
```

```
  request_id requestOrigin                     requestLabel      collection   status nrecords
2     155886           web                 steffi's test SK   SKATLAS1BE_DO approved    99872
3     155886           web                 steffi's test SK SKATLAS1BE_SUMM approved    42203
4     155886           web                 steffi's test SK      SKATLAS1PC approved    27916
                                                                  filters
2 date_type (s); project ('SKATLAS1BE_DO','SKATLAS1BE_SUMM','SKATLAS1PC')
3 date_type (s); project ('SKATLAS1BE_DO','SKATLAS1BE_SUMM','SKATLAS1PC')
4 date_type (s); project ('SKATLAS1BE_DO','SKATLAS1BE_SUMM','SKATLAS1PC')
```

```{r, eval = FALSE}
sask <- nc_data_dl(request_id = 155886, username = "USER")
```
