---
title: "Filtering data after download"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Filtering data after download"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `naturecounts` package restricts users to three types of data filters in order not to overload the NatureCounts server. However, it is fairly straightforward to apply data filters after the download. 

In this vignette we will go over several different types of filtering using the `tidyverse` collection of packages. In particular, we will make extensive use of the `filter()` function from the `dplyr` package. For more information about this function, see the chapter on [Data Transformations](https://r4ds.had.co.nz/transform.html#filter-rows-with-filter) in R for Data Science by Hadley Wickham and Garrett Grolemund.

For each example, we will show case how to filter your data either directly downloaded as a data frame, or obtained from a SQLite database.

First let's get a collection of chickadee data to use. We'll collect all public data on Carolina, black-capped and mountain chickadees (see the [vignette on species codes](species-codes.html)) for more information on how to find species codes).

## Setup

**Packages**

```{r, message = FALSE}
library(naturecounts)
library(dplyr)
library(lubridate)
```

**Downloading data directly to a Data Frame**
```{r}
chickadee_df <- nc_data_dl(species = c(14270, 14280, 14290), verbose = FALSE,
                           username = "sample", info = "vignette")
```

**Downloading data to a SQLite DataBase (`chickadee.nc`)**
```{r, echo = -1}
if(file.exists("chickadee.nc")) file.remove("chickadee.nc")
con <- nc_data_dl(species = c(14270, 14280, 14290), sql_db = "chickadee", 
                  verbose = FALSE, username = "sample", info = "vignette")
chickadee_db <- tbl(con, "naturecounts")
```

```{r, include = FALSE}
DBI::dbDisconnect(con)
```

Or, if starting a new session with an existing SQLite database, first create the connection to the database
```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = "chickadee.nc")
chickadee_db <- tbl(con, "naturecounts")
```

When using a database, you can choose to `collect()` your data into a data frame (in which case it will be identical to `chickadee_df`, but if your database is large, it will be faster to leave it as a database as long as possible.

```{r}
chickadee_df <- collect(chickadee_db)
```

We will also use a helper function to add columns for `date` and `doy` (day of year) to our data.

For data frames, you directly modify the data frame object.
```{r}
chickadee_df <- format_dates(chickadee_df)
```

For databases, you'll modify the connection, then update your table reference.

**Note:** This results in on-disk additions to the SQLite database. If you do not wish to make add fields to your SQLite database, you will have to `collect()` the data into a data frame first.

```{r}
con <- format_dates(con)
chickadee_db <- tbl(con, "naturecounts")
```


The warning reflects dates that were not valid. In this case, they were missing `survey_month` and `survey_day`.

```{r}
filter(chickadee_df, is.na(date)) %>%
  select(survey_year, survey_month, survey_day, date)
```

Unless stated otherwise, in the following examples, we demonstrate code that works for both data frames and databases.

## Categorical filters

The simplest way to filter your data is to pick out categories to include or exclude. In each case there are three steps.

1. What are the categories available?
2. Filter the categories you a) wish to keep, or b) wish to exclude
3. Confirm the results

These steps work for categories such as `collection`, `species`, `survey_year`, `iba_site`, `bcr` or any variable that can be expressed as a category.

1. What categories are available?
- `pull()` grabs the column of interest, `unique()` returns only unique values.

```{r}
chickadee_db %>%
  select(collection) %>%
  distinct()
```

2a) Filter the categories you wish to keep
```{r}
alberta_atlas <- filter(chickadee_db, 
                        collection %in% c("ABATLAS1", "ABATLAS2"))
```

2b) Filter the categories you wish to exclude with `!`
```{r}
alberta_atlas <- filter(chickadee_db, 
                        !collection %in% c("ABBIRDRECS", "RCBIOTABASE"))
```

3. Confirm
```{r}
alberta_atlas %>%
  select(collection) %>%
  distinct()
```

## Numerical filters

Another way to filter your data is to specify numerical ranges include or exclude. In each case there are the same three steps as above.

1. What is the numerical range available?
2. Filter the variable to the range you a) wish to keep, or b) wish to exclude
3. Confirm the results

These steps work for categories such as `survey_year`, `survey_month`, `survey_day`, `doy` or any variable that can be expressed as a number.

1. What categories are available?
- `pull()` grabs the column of interest, `summary()` returns a numerical summary of the values.

```{r}
chickadee_db %>%
  select(survey_year) %>%
  summarize(min = min(survey_year, na.rm = TRUE), 
            max = max(survey_year, na.rm = TRUE))
```

2a) Filter the variable to the range you wish to keep (note that the `,` is an implied `&`)
```{r}
recent <- filter(chickadee_db, survey_year >= 2015)
```

2b) Filter the variable to the range you wish to exclude with `!` (note that here we need to use the `&` explicitly)
```{r}
recent <- filter(chickadee_db, !(survey_year > 1960 & survey_year < 2015))
```

3. Confirm
```{r}
chickadee_db %>%
  select(survey_year) %>%
  summarize(min = min(survey_year, na.rm = TRUE), 
            max = max(survey_year, na.rm = TRUE))
```

## Date filters

Filtering by dates is usually slightly more complex, but fortunately, the `filter()` function automatically detects dates which means it's pretty much the same as a numerical filter.

1. What categories are available?
- `pull()` grabs the column of interest, `summary()` returns a summary of the values.

```{r}
chickadee_db %>%
  select(date) %>%
  summarize(min = min(date, na.rm = TRUE), 
            max = max(date, na.rm = TRUE))
```

2a) Filter the variable to the range you wish to keep (note that the `,` is an implied `&`)
```{r}
field_season <- filter(chickadee_db, date >= "2010-05-10", date <= "2010-09-21")
```

2b) Filter the variable to the range you wish to exclude with `!` (note that here we need to use the `&` explicitly)
```{r}
omit_field_season <- filter(chickadee_db, !(date >= "2010-05-10" & date <= "2010-09-21"))
```

3. Confirm
```{r}
field_season %>%
  select(date) %>%
  summarize(min = min(date, na.rm = TRUE), 
            max = max(date, na.rm = TRUE))
```

```{r, include = FALSE}
file.remove("chickadee.nc")
```

