---
title: "Region Codes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Region Codes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(out.width = "60%", fig.width = 6, fig.asp = 1, fig.align = "center")
```

```{r}
library(naturecounts)
```

In the main data download functions for `naturecounts`, `nc_data_dl()` and `nc_count()`, you have the option of filtering data by `region`. In this article we will explore the various ways of specifying regional filters.

For more details regarding spatial limits to regions, see the articles:

- [Mapping observations](https://birdstudiescanada.github.io/naturecounts/articles/articles/mapping-observations.html)
- [IBAs and BCRs](region-areas.html)
- [Using spatial data to filter observations](region-spatial.html)

## In short

`region` must be a named list with **one** of the following:

- `country` - Country code (e.g., `CA` for Canada)
- `statprov` - State/province code (e.g., `MB` for Manitoba)
- `subnational2` - Subnational (type 2) code (e.g., `CA.MB.07` for the Brandon Area)
- `iba` - Important Bird Areas (IBA) code (e.g., `AB001` for Beaverhill Lake in Alberta)
- `bcr` - [Bird Conservation Regions](https://www.birdscanada.org/research/gislab/index.jsp?targetpg=bcr&targetpg=bcr) (e.g., `2` for Western Alaska)
- `utm_squares` - UTM square code (e.g., `10UFE96` for a grid in Alberta)
- `bbox` - Bounding box coordinates (e.g., `c(left = -101.097223, bottom = 50.494717, right = -99.511239, top = 51.027557)` for a box containing Riding Mountain National Park in Manitoba)

To use the `region` argument:

```{r}
nc_count(region = list(statprov = "NB"))
```


## In Detail - Codes

### Country

**Search by name** (English or French)
```{r country}
search_region("États-Unis", type = "country")
```

**Browse the code list**
```{r, eval = FALSE}
meta_country_codes()
```

**Use the resulting code(s)**
```{r, eval = FALSE}
nc_count(region = list(country = c("US", "CA")))
```


### State/Province

**Search by name** (English, French, or Spanish)
```{r statprov}
search_region("Distrito de Colombia", type = "statprov")
```

**Browse the code list**
```{r, eval = FALSE}
meta_statprov_codes()
```

**Use the resulting code(s)**
```{r}
nc_count(region = list(statprov = c("DC", "MB")))
```


### Subnational regions (type 2)

**Search by name**  
Language depends on location: 
- Mexico = Spanish
- USA = English
- Quebec = French
- Rest of Canada = English

```{r subnational}
search_region("Montreal", type = "subnational2")
```

**Browse the code list**
```{r, eval = FALSE}
meta_subnational2_codes()
```

**Use the resulting code(s)**
```{r}
nc_count(region = list(subnational2 = c("CA-QC-MR", "CA-ON-TO")))
```


### Important Bird Areas (IBA)

- These are Canadian designations

**Search by name** (English or French)

```{r iba}
search_region("oak hammock", type = "iba")
```

**Browse the code list**
```{r, eval = FALSE}
meta_iba_codes()
```

**Use the resulting code(s)**
```{r}
nc_count(region = list(iba = c("MB010", "AB011")))
```

### Bird Conservation Regions (BCR)

**Search by name** (English, French, or Spanish)

```{r bcr}
search_region("rainforest", type = "bcr")
```

**Browse the code list**
```{r, eval = FALSE}
meta_bcr_codes()
```

**Use the resulting code(s)**
```{r}
nc_count(region = list(bcr = c(5, 2)))
```

### UTM Squares

**Browse the code list**
```{r, eval = FALSE}
meta_utm_squares()
```

**Use the resulting code(s)**
```{r}
nc_count(region = list(utm_squares = "18TUR35"))
```

### Bounding Box

Bounding box reflects latitude and longitude limits.

**Use the resulting code(s)**
```{r}
nc_count(region = list(bbox = c(left = -125, bottom = 45, 
                                right = -100, top = 50)))
```
