---
title: "Exploring IBAs and BCRs"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring IBAs and BCRs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(out.width = "60%", fig.width = 6, fig.asp = 1, fig.align = "center")
```


In this section we'll explore Important Bird Areas (IBAs) and Bird Conservation Regions (BCRs), so you'll have a better idea of what they represent when used as data filters.

We'll be using the [`tidyverse`](http://tidyverse.org) packages `dplyr` and `ggplot2` for data manipulation and plotting, respectively. We'll use the `sf` package for working with spatial data, and the `rnaturalearth` package to get example spatial data.

```{r, message = FALSE}
library(naturecounts)
library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(mapview)
```

First we'll get some spatial objects for our explorations from the `rnaturalearth` package. A map of North America and of Canada, both transformed from CRS EPSG code 4326 (unprojected lat/lon) to 3347 (NAD83 Statistics Canada).

```{r}
na <- ne_countries(continent = "north america", returnclass = "sf") %>%
  st_transform(3347)

canada <- ne_states(country = "canada", returnclass = "sf") %>%
  st_transform(3347)
```

### Important Bird Areas
To get a visual idea of where different Important Bird Areas are, let's plot them on our map of Canada.

First we need to grab the IBA data frame and convert it to a spatial object. Because the data contains lat/lon, we assign it to CRS 4326 for GPS lat/lon data, and then convert it to match the crs of our maps.

```{r}
iba <- meta_iba_codes() %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(3347)

ggplot() +
  theme_bw() +
  geom_sf(data = canada) +
  geom_sf(data = iba)
```

If you want to narrow in on only one province,

a) filter the IBA data frame

```{r}
iba_nwt <- filter(iba, statprov == "NT")
```

b) get the limits of the Northwest Territories

```{r}
limits <- st_bbox(filter(canada, name == "Northwest Territories"))
```

Put it all together
```{r}
ggplot() + 
  theme_bw() +
  geom_sf(data = canada) +
  geom_sf(data = iba_nwt) +
  geom_sf_text(data = iba_nwt, aes(label = iba_site), size = 3, vjust = -0.5) +
  coord_sf(xlim = limits[c("xmin", "xmax")], ylim = limits[c("ymin", "ymax")])
                               
```

For a more detailed view of IBAs, we can also grab the shape files from Birds Canada.

```{r}
iba <- st_read("https://birdmap.birdscanada.org/geoserver/atlas/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=atlas:iba&outputFormat=application%2Fjson") %>%
  st_transform(3347)

iba <- filter(iba, Status == "confirmed") # Omit historical areas

ggplot() +
  theme_bw() +
  geom_sf(data = canada, colour = "grey80") +
  geom_sf(data = iba, fill = "blue", colour = NA, show.legend = FALSE, alpha = 0.5)
```

To really explore these areas, interactive maps can be really useful. Note that you can move the map, zoom in and out, and click on specific IBAs for more information.

```{r}
mapview(iba)
```

### Bird Conservation Regions

Bird Conservation Regions are also available from Birds Canada.

```{r}
bcr <- st_read("https://birdmap.birdscanada.org/geoserver/atlas/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=atlas:bcr&outputFormat=application%2Fjson") %>%
  st_transform(3347)

ggplot() +
  theme_bw() +
  geom_sf(data = na, colour = "grey80") +
  geom_sf(data = bcr, aes(fill = bcr), show.legend = FALSE, alpha = 0.5) +
  scale_fill_viridis_c()
```

And again, interactive maps can be really useful.

```{r, fig.asp = 0.75}
mapview(bcr)
```