---
title: "Selecting columns and fields to download"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting columns and fields to download}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message = FALSE}
library(naturecounts)
library(dplyr)
```

In `naturecounts`, by default default data is downloaded with a specific set of fields/columns. However, for more advanced applications, users may wish to specify which fields/columns to return. The Bird Monitoring Data Exchange (BMDE) schema keeps track of variables used to augment observation data. There are different versions reflecting different collections of variables. 


## BMDE versions

There are several ways to get more information about these different field/columns sets.

1. You can get a list of the different versions with the `meta_bmde_versions()` function.
```{r}
meta_bmde_versions()
```

2. You can get a list of the specific fields/columns included in each version using the `meta_bmde_fields()` function
```{r}
meta_bmde_fields(version = "BMDE-BAND-2.00") %>%
  head()
```

3. You can see which BMDE versions are used by different collections by looking at the metadata associated with that collection.
```{r}
meta_collections() %>% 
  head()
```

## Specifying fields by version

By default, the `BMDE2.00-min` or `minimum` BMDE version is used when downloading data, but you can specify other versions with either the version name, or the shorthand.

```{r}
cardinals <- nc_data_dl(species = 19360, fields_set = "core", 
                        username = "sample", info = "nc_vignette")
names(cardinals)
```

## Basic fields

Note that there are extra fields/columns downloaded in every request, which are **in addition** to the BMDE fields. These are additional, basic, indexed fields used in the database. 

You can figure out exactly which fields these are, by removing all the fields from the `core` BMDE field set.

```{r}
names(cardinals)[!names(cardinals) %in% meta_bmde_fields(version = "core")$local_name]
```


## Custom fields 

The basic fields will **always** be included in any data request. However, you can choose which fields/columns to include in addition to these basic fields, by specifying `fields_set = custom`, and listing all additional `fields`.

```{r}
cardinals <- nc_data_dl(species = 19360, fields_set = "custom", 
                        fields = c("Sex", "SamplingEventIdentifier"), 
                        username = "sample", info = "nc_vignette")
head(cardinals)
```

## Adding specific fields to a BMDE version

In a more complex example, if you wished to download data with the `minimum` set of BMDE fields, but including a couple of extra fields, you could combine the core fields with the extras you want, and pass them on as a custom field set.

For example, first collect the minimum field/column names.

```{r}
my_fields <- meta_bmde_fields(version = "minimum")$local_name
my_fields
```

Then add in the extra fields.

```{r}
my_fields <- c(my_fields, "Sex", "LifeStage")
my_fields
```

Now download the data.

```{r}
cardinals <- nc_data_dl(species = 19360, fields_set = "custom", 
                        fields = my_fields, username = "sample", 
                        info = "nc_vignette")
head(cardinals)
```

Great!