---
title: "Chapter 4 - Downloading and Filtering Data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 4 - Downloading and Filtering Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Downloading and Filtering Data {#Download4}

```{r, message = FALSE}
library(naturecounts)
```

This chapter jumps right into downloading NatureCounts data, which builds directly on the skills developed in [Chapter 3](#Data3).

## Downloading NatureCounts data {#Download4.1}

To download data you need to [sign up](https://naturecounts.ca/nc/default/register.jsp) for a **free** account.

Let's try fetching observational data across multiple collections into a local data frame. To do so, we will use the `nc_data_dl()` download function. There is lots of data here, so this could take a couple minutes.

First find the species id for the target species:

```{r}
search_species("American Bittern")
```

Then use this id with `nc_data_dl()`:

```{r DataDownload, eval = FALSE}
bittern <-  nc_data_dl(species = 2490, username = "testuser", info = "tutorial example")
```

The bittern dataframe should now appear in the upper right panel of RStudio under the *Environment* tab. Notice we specified the `info` parameter which is a short description of what the data is being downloaded for. This does not need to be specified when using the `nc_count()` view function.

> The code in this Chapter will not work unless you replace `"testuser"` with your actual user name. You will be prompted to enter your password.

Next we demonstrate how to download the Ontario Whip-poor-will collection (WPWI), since it is Open Access (Level 5) and contains relatively few records (n = 3012). Click [here](https://naturecounts.ca/nc/default/datasets.jsp?code=WPWI) for more information on this dataset.

```{r OpenAccess, eval = FALSE}
WPWI <- nc_data_dl(collections = "WPWI", username = "testuser", info = "tutorial example")
```

Our data download was more specific, and way faster! This will depend on your research goals and data needs. Instead of specifying species id, we filtered our data download by using the species alpha-numeric code and the `collections` argument.

To access Level 3 or 4 collections you must [sign up](https://naturecounts.ca/nc/default/nc_access_levels.jsp) for a **free** account and request permission from the data custodian. For a complete list of datasets and access level, visit the [NatureCounts datasets](https://naturecounts.ca/nc/default/datasets.jsp) page or use the `meta_collection()` function. Here you can browse information on each dataset prior to requesting access.

To make a data request, use the NatureCounts [Download Data query tool](https://naturecounts.ca/nc/default/searchquery.jsp). For step-by-step visual instructions, we encourage you to watch: [NatureCounts: An Introductory Tutorial](link%20to%20be%20provided). You will receive an email confirmation when your request has been approved, which will contain your `request_id`. This number will be used to download your newly acquired dataset into R. There is also a build in function that allows you to check the status of your request and retrieve your `request_id`:

```{r request, eval = FALSE}
nc_requests(username = "testuser")
```

Here is sample code to download Access Level 3 or 4 data:

```{r RequestAccess, eval = FALSE}

my_data <- nc_data_dl(request_id = 000000, username = "USER",
                      info = "MY REASON")
```

**Note: This code is not functional and is here to serve as a structural example. You will need to insert your `request_id`, `username`, and `info` in the code chunk above to make this work. If you applied filters to the web portal data request (e.g., species, region, year), you will only receive the subset of the dataset you requested.**

## Applying Filters {#Download4.2}

The filters applied in [Chapter 3](#Data3) when using the `nc_count()` view function are also used for the `nc_data_dl()` download function. Again, these options include: `collections`, `project_id`, `species`, `years`, `doy` (day-of-year), `region`, and `site_type`. You may also wish to specify which fields/columns to return using the `field_set` and `fields` options in the `nc_data_dl()` function. For help with this feature, see the naturecounts article ['Selecting columns and fields to download'](https://birdstudiescanada.github.io/naturecounts/articles/selecting-fields.html).

Please review the resources provided on [filter metadata](#Data3.5) prior to proceeding.

The users can specify up to 3 filter options in the download process and should try to limit redundancies in filters. In most cases, one of these options will be `collections`, since authorization is given independently for each dataset (i.e., collection). If the user chooses to select `species`, then only collections the user has *authorization* to access will be used to retrieve species-specific data. This differs from the `nc_count()` function, which by default shows you *all* the records available in each collection.

## Examples {#Download4.3}

Here are a few examples for you to work through to become familiar with the `nc_data_dl()` function.

*Example 1*: We will use the Ontario Whip-poor-will collection (WPWI) again for this example. After reviewing the online material, [here](https://naturecounts.ca/nc/default/datasets.jsp?code=WPWI), you notice that the main survey ran from 2010 to 2012, but that additional surveys were conducted between 2009 and 2013. For your research, you only want data collected during the main survey window. Further, the survey was Ontario-wide, but most effort and records are from southern Ontario. Let's further limit the data to those collected south of the Canadian Shield, which is approximated with [BCR](http://nabci-us.org/assets/images/bcr_map2.jpg) 13.

```{r nc_dl_Filter, eval = FALSE}
WPWI_filter <- nc_data_dl(collections = "WPWI", years = c(2010, 2012),region = list(bcr = "13"), username = "testuser", info = "tutorial example")
```

You will notice that the number of records in the filtered WPWI download (`WPWI_filter` = 754) is substantially less than the number of records in the full dataset (`WPWI` = 3012) downloaded in [Chapter 3](#Data3).

*Example 2*: Rather than using BCR 13 to approximate the study area, you are interested in just looking at records collected on the [Bruce Peninsula](https://en.wikipedia.org/wiki/Bruce_Peninsula) for the full time period. You decide to use [spatial data to filter observations](https://birdstudiescanada.github.io/naturecounts/articles/region-spatial.html). Specifically, you create a [bounding box](https://birdstudiescanada.github.io/naturecounts/articles/region-codes.html), using the [Within Coordinates](https://naturecounts.ca/monitoring/georef_rec.jsp?lang=EN) tool to help you retrieve custom coordinates for your data query and/or download.

```{r nc_dl_BB, R.options=list(width=80), eval = FALSE}
WPWI_bp <- nc_data_dl(collections = "WPWI", 
                      region = list(bbox = c(left = -81.7, bottom = 44.5, 
                                             right = -80.9, top = 45.3)), 
                      username = "testuser", info = "tutorial example")

```

You will notice there are now even fewer observation records (`WPWI_bp` = 72)

Reminder that, through the naturecounts R package, data is downloaded with a specific set of fields/columns by default. However, for more advanced applications, users may wish to [specify which fields/columns to return](https://birdscanada.github.io/naturecounts/articles/selecting-fields.html).

## Exercises {#Download4.4}

Now apply your newly acquired skills!

*Exercise 1*: You are from the Northwest Territories and interested in learning more about breeding birds in your region. First, you identify the NatureCounts dataset most suitable for this exercise using the `nc_count()` function. Next, you decide to focus your download to only include Blackpoll Warbler data collected over the past 5 years (2015-2020). How many observation records did you download?

Answer: BBS50-CAN, 1756

*Exercise 2*: You are birding in the *Beaverhill Lake* Important Bird Area (iba) in May. You think you hear a Bobolink! You are curious if this species has been detected here in the [month of May](https://www.esrl.noaa.gov/gmd/grad/neubrew/Calendar.jsp). You choose to download the records you have authorization to freely access from NatureCounts. How many observation records did you download? What year are these records from?

Answer: 3, 1988 & 1987

Answers to the exercises can be found in [Chapter 7](#Ans7.2).
