---
title: "04-ElevationData"
author: "Dimitrios Markou"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 4 Digital Elevation Data

##### Author: Dimitrios Markou

The NatureCounts_SpatialData_Tutorial modules aim to help users integrate NatureCounts data with spatial and environmental layers in R and meet their conservation and research goals. In [Chapter 3](03-ClimateData.Rmd), you processed vector and raster climate data, linked them with NatureCounts data, and created maps.

Spatial data mapping can help you explore possible associations between species occurrences, climate, and other environmental variables of interest. Integrating diverse NatureCounts datasets can, therefore, help inform species distribution models, select predictor variables, and provide clear way to communicate your data.

In this tutorial, you will process Digital Terrain Models, apply crop, mask, and reclassification procedures, and extract elevation values to link with NatureCounts data.

**After downloading the necessary packages in [4.0 Learning Objectives] be sure to follow the [Preliminary Steps] section found at the end of this document before proceeding with the tutorial. It helps you preprocess the National Park boundary and NatureCounts data which are required for this lesson.**

# 4.0 Learning Objectives

By the end of **Chapter 4 - ElevationData**, users will know how to:

-   Process Digital Terrain Model (DTM)
-   Crop, mask, and reclassify raster data
-   Link NatureCounts data to elevation data for analysis

This R tutorial requires the following packages:

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(naturecounts)
library(sf)
library(terra)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggspatial)
library(mapview)
library(leaflet)
library(leaflet.extras)
library(leaflet.providers)
```

This tutorial uses the following spatial data.

1.  [Places administered by Parks Canada](https://open.canada.ca/data/en/dataset/e1f0c975-f40c-4313-9be2-beb951e35f4e/resource/0744186a-449b-4f1f-9f1e-952a94c6d5ca) - Boundary shapefiles

2.  [Forêt ouverte (lidar-derived products)](https://www.donneesquebec.ca/recherche/fr/dataset/produits-derives-de-base-du-lidar) - Digital Terrain Models (DTMs)

    **Step 1**: Navigate to the Lidar data site. Under **Données et ressources** \> **Carte interactive** \> **Forêt ouverte** select the **Téléchargement** option to explore the Lidar tiles on an interactive map (**Explorer \> Accéder**).

    **Step 2**: Using the search bar, search for Parc national de la Mauricie. Under Layers, toggle on Parc nationaux du Canada to visualize the park boundary. You can also import your own **.shp** or **.kml** file using the **Import and Export** tab.

    **Step 3**: Download the Modèle num. terrain (MNT) (Résolution spatiale 1 m) for each of the six Lidar tiles that intersect with the park (14NE, 15NO, 14SE, 15SO, 11NE, 10NO).

Save these files in you R project working directory for easy accesses during this tutorial.

```{r evel=FALSE}
getwd()
```

# 4.1 DEM, DSM & DTM

Digital elevation datasets store topographic information like elevation or slope and are a useful way to interpret landscape ecology. These datasets, i.e. Digital Elevation Models (DEMs), Digital Surface Models (DSMs), and Digital Terrain Models (DTMs) are derived through a variety of remote sensing and spatial interpolation techniques and all help describe land features.

**Digital Elevation Model** - represents the bare-Earth surface and [excludes]{.underline} all terrain vector features (i.e. streams, breaklines, and ridges), and all ground objects (power lines, buildings, trees, and vegetation).

**Digital Surface Model** - represents the heights of the Earth's surface and [includes]{.underline} all natural and artificial features or ground objects.

**Digital Terrain Model** - represents the bare-Earth surface topography and includes all terrain vector features. It does not include natural or artificial ground objects. In other words, it is a DEM that is augmented by the presence of streams, breaklines, and ridges.

Light Detection and Ranging (LiDAR) is an active remote sensing technology. It is performed using laser scanners that emit pulses of light and determine the position of target 3D objects by measuring the amount of time between pulses being emitted and received. It is a revolutionary technology that helps in the acquisition of extremely accurate land data over wide spatial and temporal scales.

Set the path to your TIF file directory.

```{r, eval = FALSE}
dir_path <- "path/to/your/directory"
```

Create a mosaic of the adjacent DTM rasters.

```{r, eval = TRUE}
# list all the TIFF files in your directory
dtm_files <- list.files(dir_path, pattern = "\\.tif$", full.names = TRUE)

# Read into a list of SpatRaster objects
dtm_list <- lapply(dtm_files, rast)

# Combine into a single SpatRaster object
dtm_mosaic <- do.call(mosaic, dtm_list)

# Print information about the stack
print(dtm_mosaic)
```

Let's check if the DTM and National Park boundary have the same crs by using the `st_crs()` function and equality operator (**==**) which will generate either TRUE or FALSE.

```{r, eval = TRUE}
st_crs(dtm_mosaic) == st_crs(mauricie_boundary)
```

To reproject the spatial data with the same CRS, we can use the `st_transform()` function.

```{r}
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(dtm_mosaic))
```

We can then use the base `plot()` function to visualize the terrain and boundary of La Mauricie National Park.

```{r regional_dtm, eval = TRUE, warning = FALSE}
# Plot the DTM mosaic
plot(dtm_mosaic, main = "DTM Mosaic with La Mauricie Boundary")

# Overlay the National Park boundary
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)
```

# 4.2 Cropping and masking

**Cropping** reduces the extent of a raster to the extent of another raster or vector.

**Masking** assigns NA values to cells of a raster not covered by a vector.

To mask a raster to a vector extent we can apply the `mask()` function from the `terra` package which uses the SpatVector format. Here, we create the mask while converting the National Park Boundary to a SpatVector using `vect()`.

```{r, warning = FALSE}
mauricie_mask <- mask(dtm_mosaic, vect(mauricie_boundary))
```

Visualize the regional and masked DTM's with a two-panel plot.

```{r masked_dem, eval = TRUE, warning = FALSE}
# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))

# Plot dtm_mosaic
plot(dtm_mosaic, main = "regional DTM")

# Overlay the National Park boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Plot mauricie_mask
plot(mauricie_mask, main = "masked DTM")

# Overlay the National Park boundary on the second plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)
```

# 4.3 Reclassification

When communicating our data, it might be useful to reclassify a range of raster values into classes. Reclassifying rasters is easily achieved using `terra` and the `classify()` function.

First, provide a reclassification table in the form of a matrix which specifies the range of raster values (columns 1 and 2) and the class ID (column 3) like so:

```{r}
dtm_rcl_matrix <- matrix(c(0, 200, 1,
                           200, 400, 2,
                           400, 1000, 3), byrow = TRUE, ncol = 3)
```

Then, apply the `classify()` function using the reclassification matrix:

```{r}
dtm_mosaic_reclass <- classify(dtm_mosaic, dtm_rcl_matrix)
```

Finally, give meaningful labels to each of the classes:

```{r, eval = TRUE}
elevation_levels <- data.frame(ID = c(1L, 2L, 3L),
                               category = c("Low elevation",
                                            "Mid elevation",
                                            "High elevation"))
levels(dtm_mosaic_reclass) <- elevation_levels
```

Visualize the regional DTM and National Park boundary using the reclassification scheme:

```{r, eval = TRUE, warning = FALSE}
# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))  # Set up a 1-row, 2-column plotting layout

# Plot the original DTM mosaic
plot(dtm_mosaic, main = "Regional DTM")

# Overlay the Mauricie boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Plot the reclassified DTM mosaic
plot(dtm_mosaic_reclass, main = "Reclassified DTM")

# Overlay the Mauricie boundary on the second plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Reset the plotting layout to the default
par(mfrow = c(1, 1))

```

Summarize the NatureCounts data for mapping and covert it to SpatVector format.

```{r}
# Group by SiteCode and summarize total_count
mauricie_site_summary <- mauricie_birds %>%
  group_by(SiteCode) %>%
  summarize(total_count = sum(ObservationCount, na.rm = TRUE))
```

```{r}
mauricie_birds_vect <- vect(mauricie_birds) # Converts to SpatVector format
```

To map the NatureCounts and DTM data:

```{r}
# Match the CRS
mauricie_site_summary <- st_transform(mauricie_site_summary, crs = st_crs(mauricie_mask))

# 

# Plot the mauricie_mask raster
plot(mauricie_mask, main = "Mauricie Mask with Site Summary Points")

# Overlay the National Park boundary in red
plot(st_geometry(mauricie_boundary), col = NA, border = "red", add = TRUE)

# Overlay mauricie_site_summary multipoints
plot(st_geometry(mauricie_site_summary), 
     add = TRUE, 
     pch = 19,       # Solid circle
     col = "blue",   # Point color
     cex = 0.5)      # Point size
```

Extract elevation values for each bird observation site and append it to `mauricie_birds`.

```{r}
mauricie_birds <- st_transform(mauricie_birds, crs(mauricie_mask)) # Match the CRS

elevation_values <- terra::extract(mauricie_mask, mauricie_birds_vect) # Extracts the elevation values for each point

elevation_values <- cbind(record_id = mauricie_birds$record_id, elevation = elevation_values[, 2]) # Creates the record_id and elevation columns 
```

Congratulations! You completed **Chapter 4 - Digital Elevation Data**. In this chapter, you successfully 1) processed raster DTM's 2) performed cropping, masking, and reclassification procedures and 3) extracted elevation data over an area of interest. In [Chapter 5](05-LandcoverData.Rmd), you can explore how to extract unique pixel values from landcover data and link it to NatureCounts data.

3)  calculated summary statistics 4) calculated NDWI and NDVI using satellite imagery

# Preliminary Steps

[La Mauricie National Park](https://parks.canada.ca/pn-np/qc/mauricie/nature) is situated in the Laurentian mountains and covers 536 km2 within the Eastern Canadian Temperate-Boreal Forest transition ecoregion. The environment is characterized by mixed forests, lakes, rivers, and hills that range from from 150 m to over 500 m in elevation. The park provides suitable habitat for a variety of wildlife including at least 215 bird species.

Read the National Park polygons into R.

```{r eval = FALSE}
#if in your working directory
national_parks<-st_read("vw_Places_Public_lieux_public_APCA.shp")

#else, specify your directory
national_parks <- st_read("path/to/your/shp")

```

Filter the national_parks dataset for La Mauricie National Park.

```{r}
View(national_parks) #to find the correct object ID

mauricie_boundary <- national_parks %>%
  filter(OBJECTID == "21")

# Drop the Z-dimension (3D component) to make it 2D
mauricie_boundary <- st_zm(mauricie_boundary, drop = TRUE, what = "ZM")
```

Write the boundary as a shapefile to your disk.

```{r, eval=FALSE, message=FALSE}
st_write(mauricie_boundary, "mauricie_boundary.shp")

# you will want the KML file for section 3.8
st_write(mauricie_boundary, "mauricie_boundary.kml", driver="KML")
```

To assess the species distribution within the National Park, download data from the [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr) which is part of a 5 year project that covers the distribution and abundance of all species breeding in the province.

Don't forget to replace `testuser` with your NatureCounts username. You will be prompted for your password.

Read in the list of species represented in NatureCounts:

```{r}
species_names <- search_species()
```

Download NatureCounts data:

```{r}
quebec_atlas <- nc_data_dl(collections = "QCATLAS2PC", username = "testuser", info = "spatial_data_tutorial", timeout = 500)
```

eBird has the greatest number of provincial bird records, however, this collection comprise data of Access Level 4. If you wish to access this collection you must sign up for a free account and [make a data request](https://naturecounts.ca/nc/default/explore.jsp#download). Otherwise, you can carry forward with the tutorial without these data and skip this code chunk.

```{r, eval = FALSE}
quebec_atlas <- nc_data_dl(collections = c("QCATLAS2PC", "EBIRD-CA-QC"), username = "testuser", info = "spatial_data_tutorial")
```

To create date and doy columns and ensure that the ObservationCount column is in the correct numeric format we can apply the `format_dates()` and `mutate()` functions. We will also filter the dataset to exclude rows with missing coordinates.

```{r}
quebec_atlas <- quebec_atlas %>%
  format_dates() %>%  # create the date and doy columns 
  mutate(ObservationCount = as.numeric(ObservationCount)) %>%  # convert to numeric format
  filter(!is.na(longitude) & !is.na(latitude))  # remove rows with missing coordinates
```

To convert the NatureCounts data to a spatial object and transform its crs to match the National Park boundary we can use the `st_as_sf()` and `st_transform()` functions, respectively.

```{r}
quebec_atlas_sf <- sf::st_as_sf(quebec_atlas,
                        coords = c("longitude", "latitude"), crs = 4326) # converts the quebec_atlas data to an sf object

mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(quebec_atlas_sf)) # match the CRS
```

Clip the NatureCounts data to the National Park boundary using `st_intersection()`.

```{r, warning=FALSE}
mauricie_birds <- sf::st_intersection(quebec_atlas_sf, mauricie_boundary)
```

Append the species names to the clipped NatureCounts dataset based on `species_id` code.

```{r}
mauricie_birds <- mauricie_birds %>%
  left_join(species_names, by = "species_id")
```

Tidyverse functions can help us summarize our data in a variety of ways. For example, if we wanted to determine the annual bird count for each year across all sites, we could use `mutate()` and `lubridate()` to extract the survey **year** from the **date** column. The `group_by()` function can then be used to group the observations by year, and `summarise()` can help calculate and create the **annual_count** column. Here, we ensure that the **ObservationCount** is in the correct format by applying `as.numeric()`.

```{r}
mauricie_birds_summary <- mauricie_birds %>%
  mutate(year = lubridate::year(date)) %>% # extracts the survey year using lubridate
  group_by(year) %>%
  summarise(annual_count = sum(as.numeric(ObservationCount), na.rm = TRUE)) %>% # calculates the annual_count
  filter(!is.na(year))  # remove rows with missing year

mauricie_birds_summary
```

If you wanted to summarize total count for each species at each site (i.e., atlas block within the park boundary), respectively, you could adjust the pipe like so using `pivot_wider()`.

```{r}
mauricie_species_summary <- mauricie_birds %>%
  st_drop_geometry() %>%  # drop the geometry column
  group_by(species_id, Locality) %>%
  summarise(total_count = sum(as.numeric(ObservationCount), na.rm = TRUE)) %>% # calculates the total_count column
  pivot_wider(names_from = species_id, # populates the column names with each species_id code
              values_from = total_count, # populates each cell with total_count
              values_fill = list(total_count = 0)) %>% # missing values are zero-filled
  group_by(Locality)

mauricie_species_summary
```
