---
title: "Mapping Observations"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping Observations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this article we'll walk through how to create various types of maps of the observations downloaded with `naturecounts` to get a sense of the spatial distribution.

> The following examples use the "testuser" user which is not available to you. 
> You can quickly [sign up for a free account](https://naturecounts.ca/nc/default/register.jsp)
> of your own to access and play around with these examples. Simply replace
> `testuser` with your own username.

## Setup

To do so we're going to use the following packages:

```{r, message = FALSE}
library(naturecounts)
library(sf)
library(rnaturalearth)
library(ggplot2)
library(ggspatial)
library(dplyr)
library(tidyr)
library(mapview)
```

First we'll use download some data:

```{r}
house_finches <- nc_data_dl(species = 20350, region = list(statprov = "AB"), 
                            username = "testuser", info = "nc_tutorial")
head(house_finches)
```

## Simple Maps
Here we're going to take a quick look at the spatial distribution using `ggplot2` 
and `ggspatial` to get some basemaps.

First let's get an idea of how many distinct points there are (often multiple observations are recorded for the same location).
```{r}
nrow(house_finches)

select(house_finches, longitude, latitude) |>
  distinct() |>
  nrow()
```

So we have `r nrow(unique(house_finches[, c("longitude", "latitude")]))` sites for `r nrow(house_finches)` observations.

Next let's convert our data to spatial data so we can plot it spatially (i.e. make a map!).
Note that we're using CRS EPSG code of 4326 because that's reflects unprojected, GPS data in lat/lon. 
First we omit `NA`s because `sf` data frames cannot have missing locations.
```{r}
house_finches <- drop_na(house_finches, "longitude", "latitude")
house_finches_sf <- st_as_sf(house_finches, 
                             coords = c("longitude", "latitude"), crs = 4326)
```

Now we're ready to make a map of the distribution of observations. 

We'll use a baselayer from OpenStreetMap and then add our observations.
```{r}
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = house_finches_sf) +
  labs(caption = "Copyright OpenStreetMap contributors")
```

Let's count our observations for each site.
```{r}
cnt <- house_finches_sf |>
  count(geometry)

ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = cnt, aes(size = n)) +
  labs(caption = "Copyright OpenStreetMap contributors")
```

## Interactive Maps
If we want to get fancy we can also create interactive maps using the `mapview` packages (see also the [`leaflet` for R package](https://rstudio.github.io/leaflet/)).
```{r, eval = FALSE}
mapview(house_finches_sf, zcol = "survey_year", at = seq(1965, 2005, by = 10),
        map.types = "Esri.WorldImagery")
```

## More Complex Maps

For more complex, or detailed maps, we can use a variety of spatial data files to layer our data over maps of the area.

For this we'll get some outlines of Canada and it's Provinces and Territories from `rnaturalearth`.

```{r}
canada <- ne_states(country = "canada", returnclass = "sf") |>
  st_transform(3347)

ggplot() +
  theme_bw() +
  geom_sf(data = canada)
```


Let's add our observations (note that the data are transformed to match the projection of the first layer, here the `canada` data).
```{r}
ggplot() +
  theme_bw() +
  geom_sf(data = canada) +
  geom_sf(data = house_finches_sf, size = 0.5)
```

We can also focus on Alberta

```{r}
ab <- filter(canada, name == "Alberta")

ggplot() +
  theme_bw() +
  geom_sf(data = ab) +
  geom_sf(data = house_finches_sf, size = 0.5)
```

Perhaps we should see how these observations are distributed among Alberta's 
Bird Conservation Regions (BCRs). 

First we'll grab the BCR shape file from [Birds Canada](https://www.birdscanada.org/bird-science/nabci-bird-conservation-regions)[^1].

[^1]: Bird Studies Canada and NABCI.  2014.  Bird Conservation Regions.  Published by Bird Studies Canada on behalf of the North American Bird Conservation Initiative.   https://birdscanada.org/bird-science/nabci-bird-conservation-regions  Accessed: `r Sys.Date()`

We'll download the zip file and extracted to a folder easier to work with. 
Note that these spatial data are in the GDB format, so we're working with a
folder named `XXX.gdb`.
```{r}
download.file("https://services1.arcgis.com/d5M16PKlQTMEVyua/arcgis/rest/services/BCR_terrestrial_master/FeatureServer/replicafilescache/BCR_terrestrial_master_-8688020410858769740.zip", "BCR_terrestrial_master.zip")
unzip("BCR_terrestrial_master.zip", junkpaths = TRUE, exdir = "BCR_terrestrial_master.gdb")
```


```{r}
bcr <- st_read("BCR_terrestrial_master.gdb") |>
  st_transform(3347) |>
  st_intersection(ab)
```

Add this layer to our plot.

```{r}
ggplot() +
  theme_bw() +
  geom_sf(data = ab) +
  geom_sf(data = bcr, aes(fill = factor(bcr_label)), alpha = 0.5, colour = NA) +
  geom_sf(data = house_finches_sf, size = 0.5) +
  scale_fill_viridis_d(name = "BCR")
```

Some of the border cases are a bit hard to separate visually.
To solve this problem, we can merge our observations with the BCRs and plot the
observations by BCR as well.

First we'll transform our observation data to match the CRS of `bcr`, then we'll join the BCR information to our observations, based on whether the observations overlap a BCR polygon (by default this is a left join).
Now we have a BCR designation for each observation.

```{r}
house_finches_sf <- house_finches_sf |>
  st_transform(st_crs(bcr)) |>
  st_join(bcr)
```

Those border cases are now resolved.
```{r}
ggplot() +
  theme_bw() +
  geom_sf(data = ab) +
  geom_sf(data = bcr, aes(fill = factor(bcr_label)), alpha = 0.25, colour = NA) +
  geom_sf(data = house_finches_sf, aes(fill = factor(bcr_label)), shape = 21) +
  scale_fill_viridis_d(name = "BCR")
```

We might also be interested in observations over time.

First we'll bin our yearly observations
```{r}
house_finches_sf <- mutate(house_finches_sf, 
                           years = cut(survey_year, 
                                       breaks = seq(1960, 2010, 10), 
                                       labels = seq(1960, 2000, 10), right = FALSE))
```

We'll also want to see how many sample years there are per decade.
```{r}
years <- house_finches_sf |>
  group_by(years) |>
  summarize(n = length(unique(survey_year)), .groups = "drop")
```


Now we can see how House Finch observations change over the years

```{r}
ggplot() +
  theme_bw() +
  geom_sf(data = ab) +
  geom_sf(data = bcr, aes(fill = factor(bcr_label)), alpha = 0.5, colour = NA) +
  geom_sf(data = house_finches_sf, size = 0.5) +
  scale_fill_viridis_d(name = "BCR") +
  geom_sf_text(data = years, x = 4427134, y = 2965275, hjust = 0, vjust = 1, 
               aes(label = paste0("n = ", n))) +
  facet_wrap(~years)
```

## Presence/Absence

We can also use some of the `naturecounts` helper functions to create presence/absence maps. 

Here we download data from the `RCBIOTABASE` collection, make sure to keep only observations where all species and the location were reported, create a new `presence` column which is either TRUE, FALSE, or NA for each sampling event. Finally we use the `format_zero_fill()` function to fill in sampling events where cardinals (`species_id` 19360) were not detected (presence would then be 0).

```{r}
cardinals <- nc_data_dl(collection = "RCBIOTABASE", username = "testuser", 
                        info = "nc_tutorial")

cardinals_zf <- cardinals |>
  filter(AllSpeciesReported == "Yes", !is.na(latitude), !is.na(longitude)) |>
  group_by(species_id, AllSpeciesReported, SamplingEventIdentifier) |>
  summarize(presence = sum(as.numeric(ObservationCount)) > 0, .groups = "drop") |>
  format_zero_fill(species = 19360, by = "SamplingEventIdentifier",
                   fill = "presence")
```


A bit convoluted, but here we'll grab coordinates for each Sampling Event.
This is only necessary if there are errors (worth reporting!) where there are
more than one lat/lon combo for each sampling event.

```{r}
coords <- cardinals |>
  select("SamplingEventIdentifier", "latitude", "longitude") |>
  group_by(SamplingEventIdentifier, latitude, longitude) |>
  mutate(n = n()) |>  # Count number of unique coordinates
  group_by(SamplingEventIdentifier) |>
  slice_max(n, with_ties = FALSE) |> # Grab the most common coordinates for each event
  select(-"n")

cardinals_zf <- left_join(cardinals_zf, coords, by = "SamplingEventIdentifier")

head(cardinals_zf)
```

Now that we have our presence/absence data for cardinals, we can create a map.
```{r}
cnt <- st_as_sf(cardinals_zf, coords = c("longitude", "latitude"), crs = 4326) |>
  group_by(presence) |>
  count(geometry)

ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = cnt, aes(size = n, colour = factor(presence)), alpha = 0.5) +
  scale_colour_manual(name = "Presence/Absence", values = c("#31688E", "#440154"), 
                      labels = c("1" = "Present", "0" = "Absent")) +
  scale_size_continuous(name = "Number of Sampling Events", range = c(1, 20)) +
  labs(caption = "Copyright OpenStreetMap contributors",
       title = "Presence/Absence of Cardinals in the RCBIOTABASE collection")
```

## Clean up the mapping files if you no longer need them

```{r}
unlink("alberta_parks/", recursive = TRUE)
```



## See Also

- [Using spatial data to filter observations](region-spatial.html)
- [Exploring regional filters](../region-areas.html)
